Sources of Measurement Bias on Modern CPUs

### 2013-12-20 paper.tex
Backup

@@

%\pgfplotstableread{resources/motivation.dat}{\motivationtable}
%\pgfplotstabletypeset[string type]{\motivationtable}

%\begin{figure}[t]
%  \caption{Ticz figure}
%  \begin{tikzpicture}
    %\draw circle (5);
%    \begin{axis}[
        %width = \columnwidth,
        %scale mode = stretch to fill,
%        font = \small,
        %scaled ticks = false,
%        xlabel=Characters added to environment,
        %ymin = 600000, ymax = 900000,
%        grid = major,
 %       cycle list name = exotic
 %%     ]
 %     \addplot[smooth] table[y = cycles:u] \motivationtable ;
%      \addlegendentry{cycles:u} ;
%    \end{axis}
%  \end{tikzpicture}
%\end{figure}

In this paper, we unveil some of the actual causes of bias that can be experienced by altering memory layout.
Both the size of environment variables and link ordering ultimately has an effect on memory layout of running processes. 
We will not look at things like cache efficiency, which often can be the reason for bad performance under certain memory contexts. 
Instead, we look at effects from less known optimizations in the out-of-order execution engine and instructions fetch pipeline. 
Because bias effects are intrinsically connected to various hardware features, this paper is focued on the Intel® Core™ i7 “Haswell” architecture specifically. 
However, the bias sources we present has also been shown to behave similarly on previous Intel architectures as well.
With a better understanding of intricate properties of the CPU, we show how one is able to not only predict, but also explicitly optimize for beneficial memory layouts.

@@

\pgfplotstableread{resources/stack-offset.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \small,
        xlabel=Bytes added to environment,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        legend pos = outer north east,
        cycle list name = exotic
      ]
      \draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

The stack is normally placed close to the upper address 0x7fff'ffffffff.\footnote{Modern processors do not actually use the full 64 bit space, only the low order 54 bits are used for addressing memory}
Environment variables are allocated before the stack, offsetting all call frames and addresses of stack allocated variables at run time.
There is still some level of stack alignment happening independent of environment offsets. For example, the compiler will often enforce alignment to some number of zero bits.
As long as this number is less than 12, environment size can impact addresses of automatic variables $\bmod 4096$.

@@

Note that the initial position of stack will not depend on environment size this way when address randomization (ASLR) is enabled.
\begin{figure}
  \begin{tikzpicture}
    \draw [fill=white, thick] (0, 0) rectangle (2.6, 6) ;
    \draw [fill=white, thin] (0.1, 5.3) rectangle (2.5, 5.9) ;
    \draw [fill=white, thin] (0.1, 4.0) rectangle (2.5, 5.0) ;
    \draw [fill=white, thin] (0.1, 1.6) rectangle (2.5, 2.5) ;
    \draw [fill=white, thin] (0.1, 0.6) rectangle (2.5, 1.5) ;
    \draw [fill=white, thin] (0.1, 0.1) rectangle (2.5, 0.5) ;

    \node [below right] at (2.6, 6.0) {0x7fff'ffffffff} ;
    \node [above right] at (2.6, 0.5) {0x400000} ;
    \node [above right] at (2.6, 0.0) {0x0} ;

    \node [below right] at (0.1, 5.9) {environment} ;
    \node [below right] at (0.1, 5.0) {stack} ;
    \node [below right] at (0.1, 2.5) {data} ;
    \node [below right] at (0.1, 1.5) {text} ;

    \draw [<->, line width=1, color=black] (1.3, 5.3) -- (1.3, 5.0) ;

  \end{tikzpicture}
  \caption{Stack position is offset by size of environment variables}
  \label{fig:stack}
\end{figure}

@@

To illustrate how this can cause bias, we revisit the example first presented in \cite{Mytkowicz:2009:WrongData}.
Their program contains four variables, two of which are stack allocated and i, j and k that are statically allocated.

@@

\begin{small}
\begin{table}
  \caption{Properties of the LSD}
  \begin{tabular}{@{} l r @{}}
    \toprule
      UnHalted Core Cycles & 1 \\
      LD\_BLOCKS\_PARTIAL.\_ADDRESS\_ALIAS & 0.89 \\
      CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING & 0.97 \\
      RESOURCE\_STALLS.ANY & 0.59 \\
    \bottomrule
  \end{tabular}
\end{table}
\end{small}

\subsection{Heap Address Aliasing}
As an example of a “sliding window” program working on pairs of memory buffers, consider the function shown in Listing [alg:convolution]. 
It computes the convolution between an input array and a fixed kernel, writing the result to another array (endpoints skipped for simplicity).

\begin{lstlisting}[language=C]
static float k[5]={0.1,0.25,0.3,0.25,0.1};
void convolve(int size,float *in,float *out)
{
  int i, j;
  for (i=2;i<size-2; ++i) {
    out[i] = 0;
    for (j=0;j<5; ++j)
      out[i]+=in[i-2+j]*k[j];}
}
int main() {
  float *in=malloc(N*sizeof(float));
  float *out=malloc(N*sizeof(float));
  convolve(N, in, out);
  return 0;
}
\end{lstlisting}

>>>>

\usetikzlibrary{patterns,shapes,positioning,calc}

@@

Both the size of environment variables and link ordering ultimately has an effect on memory layout of running processes. 
We will not look at things like cache efficiency, which often can be the reason for bad performance under certain memory contexts. 
Instead, we look at effects from less known optimizations in the out-of-order execution engine and instructions fetch pipeline. 
Because bias effects are intrinsically connected to various hardware features, this paper is focued on the Intel® Core™ i7 “Haswell” architecture specifically. 
However, the bias sources we present has also been shown to behave similarly on previous Intel architectures as well.
With a better understanding of intricate properties of the CPU, we show how one is able to not only predict, but also explicitly optimize for beneficial memory layouts.

@@

\pgfplotstableread{resources/stack-offset.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \small,
        xlabel=Bytes added to environment,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        legend pos = outer north east,
        cycle list name = exotic
      ]
      %\draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

The stack is normally placed close to the upper address 0x7fff'ffffffff.\footnote{Modern processors do not actually use the full 64 bit space, only the low order 47 bits are used for addressing memory}
Environment variables and program arguments are allocated in the stack section, before the first call frame.
Changing environment variables will therefore offset the addresses of all stack allocated data.
There is still some level of stack alignment happening, independent of environment offsets. 
The compiler can enforce alignment to some number of zero bits.
As long as this number is less than 12, environment size can impact addresses of automatic variables $\bmod 4096$.

@@

Note that the initial position of stack will not depend on environment size this way when address randomization (ASLR) is enabled. (TODO: Does it offset from page boundary? That would be good, same alias behavior :)

\begin{figure}
  \begin{tikzpicture}[font=\footnotesize]

    % See page 453 in pgf manual
    \node [
      rectangle split, rectangle split parts=9, 
      rectangle split part fill={white, lightgray, white, lightgray, white, lightgray, white},
      draw, anchor=center, text width=2cm
    ] (m)  
      {
        \nodepart{one}
          stack
        \nodepart{three}
          mmap area
        \nodepart{five}
          heap
        \nodepart{seven}
          bss
        \nodepart{eight}
          data
        \nodepart{nine}
          text
      };

    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.six split west) 
      node [black, midway, xshift=-1.5cm, text width=2.0cm] 
        { Program code and static data } ;

    \node[below right] at (m.north east)
      { \scriptsize{0x7fff'ffffffff} } ;

    \node[above right] at (m.south east)
      { \scriptsize{0x400000} } ;

  \end{tikzpicture}
  \caption{Important sections of virtual memory on a 64 bit process}
  \label{fig:stack}
\end{figure}

@@

To illustrate how this can cause bias, we revisit the example first presented in \cite{Mytkowicz:2009:WrongData}.
Their program contains five variables, two of which are stack allocated and i, j and k that are statically allocated.

@@

\subsection{Heap Address Aliasing}
Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
The previous section looked specifically at an example of collision between stack variables and static data.
Another scenario to consider is collisions in heap allocated memory.
In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
A typical malloc implementation uses two different strategies for how to allocate memory.
The \emph{heap} area in virtual memory is used for smaller allocations.
Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
For larger allocations, it is typically faster to call mmap instead of increasing the ``normal'' heap.
The heap starts at a low address close to the data section and grows towards higher addresses. 
Mmap allocates much closer to the stack.
While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, mmap has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{Libc's version of malloc adds 16 bytes of medatada at the beginning, thus every mmap'ed address ends with 0x010}

Consider the function shown in Listing [alg:convolution].
It computes the convolution between an input array and a fixed five-element kernel, writing the result to another array.
For simplicity, endpoints are not handled.

\begin{lstlisting}[language=C]
static float k[5]={0.1,0.25,0.3,0.25,0.1};
void conv(int n, float *in, float *out)
{
  int i, j;
  for (i = 2; i < n-2; ++i) {
    out[i] = 0;
    for (j = 0; j < 5; ++j)
      out[i] += in[i-2+j] * k[j];
  }
}
\end{lstlisting}

\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
\begin{figure}[t]
  \caption{Offseting arrays}
  \begin{tikzpicture}
    \begin{axis}[
        font = \small,
        xlabel=Number of float (4B) offset,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convolutiontable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convolutiontable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure}
